Algebraic File Synchronization: Adequacy and Completeness

### 2016-10-16 p2/def-commands.tex
Starting with the definition of domains
New ordering method; hyperlinks

Notice, however, that the commands listed above also encode some information about 
their input, the state of the filesystem
before the command is applied. In particular, $create(n,\valv)$ requires that there are no files
or directories at $n$, while $edit(n,\valv)$ and $remove(n)$ require the opposite.
This creates an arbitrary asymmetry where
there is now more information available about their output than about their input.
As, based on the above, we expect that encoding more information in the commands
results in a model with greater predictive powers,
and in order to resolve this asymmetry, 
we propose a set of commands that encode
the type of the original value $\FS(n)$ as well.
(Some real-life filesystem commands like $rmdir$ do this already.)
Again, as the success of failure of commands depends on types of values,
it is not necessary to encode the actual input value.
For example, the behavior of two commands that differ only in what the original file contains
that they overwrite is never going to be different.

@@

So that we could reason about sequences of commands that break every filesystem, 
we also introduce the command $\cbrk$ which is not defined anywhere,
that is, which simply breaks every filesystem.

@@

\begin{mydef}[Filesystem commands]
In other words, we model filesystem commands as partial endofunctions on filesystems,
that is, functions that map filesystems to filesystems, but which
are not defined everywhere:
\[ \alpha: \setf \nrightarrow \setf \]
A command $\alpha$ is not defined on filesystems that it would break.
At times in these cases we write that the command returns the bottom element, $\fsbroken$.
We write $\cbrk$ for the empty partial function which is not defined anywhere.

>>>>

Notice, however, that the commands listed above also encode some information about 
their input, the state of the filesystem
before the command is applied. In particular, $create(n,\valv)$ requires that there are no files
or directories at $n$, while $edit(n,\valv)$ and $remove(n)$ require the opposite.
This creates an arbitrary asymmetry where
there is now more information available about their output than about their input.
As, based on the above, we expect that encoding more information in the commands
results in a model with greater predictive powers,
and in order to resolve this asymmetry, 
we propose a set of commands that encode
the type of the original value $\FS(n)$ as well.
(Some real-life filesystem commands like $rmdir$ do this already.)
Again, as the success of failure of commands depends on types of values,
it is not necessary to encode the actual input value.

@@

So that we could reason about sequences of commands that break every filesystem, 
we also introduce the command $\cbrk$ which is not defined anywhere,
that is, which breaks every filesystem.

@@

\begin{mydef}[Filesystem commands]
In other words, we model filesystem commands as partial endofunctions on filesystems,
that is, functions that map filesystems to filesystems, but which
are not defined everywhere:
\[ \alpha: \setfs \nrightarrow \setfs \]
A command $\alpha$ is not defined on filesystems that it would break.
At times in these cases we write that the command returns the bottom element, $\fsbroken$.
We write $\cbrk$ for the empty partial function which is not defined anywhere.

### 2016-10-16 p2/def-filesystem.tex
Improved notation
New ordering method; hyperlinks

We model a working filesystem
using a function $\FS$ with a set of nodes (filesystem paths) $\setn$ as its domain,
and a set of possible contents or values $\setv$ as its codomain.
\begin{mydef}[Filesystems, $\setn$, $\setv$ and $\setf$]
\[ \FS: \setn \rightarrow \setv \]
In our model, $\setn$ 
serves as a namespace or ``skeleton'' for the filesystem.
It contains all possible nodes, including the ones 
where the file system contains no file or directory,
where in our model $\FS$ has a special value, $\empt\in\setv$.
We write $\setfs$ for the set of filesystems.
\end{mydef}

@@

\begin{mydef}[$n\descendant m$]
On $\setn$ the \emph{ancestor / descendant} relation is the
strict partial ordering determined by the $\parent$ function.
We write $n\descendant m$ if $n$ is the ancestor of $m$,
that is, $n=\parent^i(m)$ for some integer $i\ge 1$.
% or the path name of $n$ is an initial segment of that of $m$. 
We write $n\descendantEq m$ if $n\descendant m$ or $n=m$.
\end{mydef}

@@

\begin{mydef}[Equivalence by type: $\typeeq$]
We write $v_1\typeeq v_2$ for $v_1,v_2\in\setv$ if $v_1$ and $v_2$ are both files,
are both directories, or are both $\empt$.
% In other words, if $\setf$ is the set of file values and $\setd$ is the set of directories, then
% with $\{\empt\}$ they form a partition of $\setv$:
% \[ \setv/{\typeeq} = \{\setd,\setf,\{\empt\}\}. \]
\end{mydef}

>>>>

We model a working filesystem
using a function $\FS$ with a set of nodes (filesystem paths) $\setn$ as its domain,
and a set of possible contents or values $\setv$ as its codomain.
\begin{mydef}[Filesystems, $\setn$, $\setv$, $\setfs$ and $\empt$]
\[ \FS: \setn \rightarrow \setv \]
In our model, $\setn$ 
serves as a namespace or ``skeleton'' for the filesystem.
It contains all possible nodes, including the ones 
where the file system contains no file or directory,
where in our model $\FS$ has a special value, $\empt\in\setv$.
We write $\setfs$ for the set of filesystems.
\end{mydef}

@@

\begin{mydef}[$\descendant$, $\descendantEq$]
On $\setn$ the \emph{ancestor / descendant} relation $\descendant$ is the
strict partial ordering determined by the $\parent$ function.
We write $n\descendant m$ if $n$ is the ancestor of $m$,
that is, $n=\parent^i(m)$ for some integer $i\ge 1$.
% or the path name of $n$ is an initial segment of that of $m$. 
We write $n\descendantEq m$ if $n\descendant m$ or $n=m$.
\end{mydef}

@@

\begin{mydef}[Equivalence by type: $\typeeq$]
We write $v_1\typeeq v_2$ for $v_1,v_2\in\setv$ iff $v_1$ and $v_2$ are both files,
are both directories, or are both $\empt$.
% In other words, if $\setf$ is the set of file values and $\setd$ is the set of directories, then
% with $\{\empt\}$ they form a partition of $\setv$:
% \[ \setv/{\typeeq} = \{\setd,\setf,\{\empt\}\}. \]
\end{mydef}

### 2016-10-16 p2/ordering.tex
Improved notation
New ordering method; hyperlinks

In this section we describe an algorithm with which all possible orderings of a
simple set of commands can be generated, but before doing so, let us consider the following lemma.

@@

Let $A$ be a sequence that satisfies the conditions;
we therefore know that it contains a command $\cxynv$ on $n$
and another command $\czwnnv$ on $\nn$.
We use a proof by contradiction and assume that there are no commands on $\parent(n)$ in $A$.
Next, we create a new sequence $A'\equiv A$ in which $\cxynv$ and $\czwnnv$ are next to each other.
If they are already next to each other in $A$, there is nothing to do.
Otherwise, consider the command next to $\cxynv$ in the direction where $\czwnnv$ is.
Let this command be $\cqrmv$.
If $\czwnnv$ is to the right, then $A$ looks like the following:
\[ A = \cdots\cc\cxynv\cc\cqrmv\cc\cdots\cc\czwnnv\cc\cdots \]
If $n\unrel m$, then swap $\cxynv$ and $\cqrmv$. Based on \cref{ax_separate_commute} we know that the new
sequence is equivalent to $A$.
Otherwise, we know $m\neq\parent(n)$ as there are no commands on $\parent(n)$, and so
from \cref{ax_distantrel_breaks} we get $A\equiv\cbrk$ which contradicts our assumptions.
(Note that $A$ cannot contains assertion commands.)
By repeating this step we can therefore convert $A$ into $A'$ where $\cxynv$ and $\czwnnv$ are neighboring commands.
However, then \cref{ax_distantrel_breaks} applies to the sequence and therefore $A\equiv A'\equiv\cbrk$ which
is again a contradiction.
\end{proof}

@@

\noindent
We can now define our algorithm to generate orders of sets of commands.

\begin{mydef}[$\ordersetsign$]
We use $\orderset{U}$ to denote the set of sequences 
generated by the algorithm / (all possible orderings)
from the commands in the simple-? set or sequence $U$.
\end{mydef}

Let $U$ be a simple set of commands
that we know can be applied to at least one filesystem in some order without breaking it.
Then we can split $U$ into a maximum number of disjoint subsets
(components) so that the nodes of two commands from two different subsets are incomparable.
These will simply be the components of the subset of the node forest that contains 
the nodes affected by $U$.

From \cref{ax_separate_commute} we know that commands from different 
components can be applied in any order.
Therefore once we have ordered the commands inside the components, we generate
all permutations of the components themselves, including the ones
in which commands from separate components mix and overlap
(but where the possible orders determined for commands inside the components are respected).

Next, consider the ordering inside a component.

This is not a problem in the case of components containing a single command only.
(Note that a file-to-file command ($\cff$) can only appear in a component of one.)

In each component that is larger than one, 
we know that all nodes are related, and from \cref{connected_changes}
we also know that it must contain commands on nodes in between the related nodes.
Therefore the nodes
form a rooted ordered tree,
and the pairs of commands connected by the edges in the tree 
must be construction or destruction pairs.
(Otherwise, there would be no ordering of the component that is not equivalent to $\cbrk$
and via \cref{ax_separate_commute} there would be no ordering of the whole set, either.)

To order the commands in the component,
select the command with the topmost node, so that there is no command
on its parent node (if there is one). 
If the command removes content at this node (it is of the form $\caa{X}{\ccharb}$), 
then it must follow all other commands in the component, and
all parent--child pairs in the whole component must be destruction pairs.
Otherwise, the command with the topmost node needs to precede the other commands,
and the component must be made of construction pairs.
These follow from the patterns of construction and destruction pairs, which
at no point can mix without causing any ordering to break all filesystems.

Then move the command with the topmost node to the beginning or the end of
the sequence we are creating out of the component,
and recursively start the whole algorithm again on the commands in the component that remain,
potentially splitting them into sub-components.
The result of this process is a set of possible orders for each component,
which are then combined into a set of possible orders for the whole set of commands.

\noindent As a specific case, we can regard a simple sequence $S$
as a set, and use the above algorithm to generate all possible orderings
of the commands it contains.
\Cref{simple_reorder_equiv} is
an important property of simple sequences: namely, that
all of their valid reorderings are equivalent,
and all equivalent sequences are mere reorderings.
To prove this, first we prove the following:

@@

\begin{mylem}\label{simple_reorder_equiv}
For a simple sequence $S\nequiv\cbrk$,
\[ \orderset{S} = [S]_\equiv, \]
where $[S]_\equiv$ denotes the set of simple sequences equivalent to $S$.
\end{mylem}
\begin{proof}
First, as $S$ only has at most one command on each node,
any sequence in $\orderset{S}$ either breaks a filesystem,
or it leaves it in the same state as $S$.

Second, based on the algorithm we also know that $\orderset{S}$ contains
all possible orderings of $S$ as the only ordering between commands
it prescribes are when the commands in the reverse order would break every filesystem.

>>>>

In this section we describe a method with which all possible orderings of a
simple set of commands can be generated, but before doing so, let us consider the following lemma.

@@

Let $A$ be a sequence that satisfies the conditions;
we therefore know that it contains a command $\cxynv$ on $n$
and another command $\czwnnv$ on $\nn$.
We use a proof by contradiction and assume that there are no commands on $\parent(n)$ in $A$.
Next, we create a new sequence $A'\equiv A$ in which $\cxynv$ and $\czwnnv$ are next to each other.
If they are already next to each other in $A$, there is nothing to do.
Otherwise, consider the command next to $\cxynv$ in the direction where $\czwnnv$ is.
Let this command be $\cqrmv$.
If $\czwnnv$ is to the right, then $A$ looks like the following:
\[ A = \cdots\cc\cxynv\cc\cqrmv\cc\cdots\cc\czwnnv\cc\cdots \]
If $n\unrel m$, then swap $\cxynv$ and $\cqrmv$. Based on \cref{ax_separate_commute} we know that the new
sequence is equivalent to $A$.
Otherwise, we know $m\neq\parent(n)$ as there are no commands on $\parent(n)$, and so
from \cref{ax_distantrel_breaks} we get $A\equiv\cbrk$ which contradicts our assumptions.
(Note that $A$ cannot contain assertion commands.)
By repeating this step we can therefore convert $A$ into $A'$ where $\cxynv$ and $\czwnnv$ are neighboring commands.
However, then \cref{ax_distantrel_breaks} applies to the sequence and therefore $A\equiv A'\equiv\cbrk$ which
is again a contradiction.
\end{proof}

@@

\noindent
We can now define the method to generate possible orders of sets of commands.
What we expect is that if the commands can be applied to a filesystem
in some order without breaking it, then the method will yield all possible orders
in which they can be applied.

\begin{mydef}[$\ordersetsign$]
For a simple set or sequence $U\nequiv\cbrk$,
we use $\orderset{U}$ to denote the set of sequences
that are all permutations of the commands in $U$
for which the following holds:
for any two commands $\cxynv$ and $\czwmv$ in a sequence in $\orderset{U}$
where $\cxynv$ comes earlier in the sequence than $\czwmv$,
\begin{itemize}
\item if $n=\parent(m)$, then $y=\vald$
\item if $\parent(n)=m$, then $y=\empt$.
\end{itemize}
\end{mydef}

The principal ideas behind this algorithm are 
\cref{ax_separate_commute}, from which we know that commands on incomparable
nodes can be applied in any order, so we only need to focus on the order of commands
with comparable nodes;
\cref{connected_changes}, from which we know that if $U$ contains commands on two comparable
nodes, then it contains commands on all nodes in between, so it is enough to specify
the order of commands on parent and child nodes;
and finally, \cref{ax_directchild_breaks,ax_directparent_breaks}, from which we know
that commands on parent--child pairs must be construction or destruction pairs.
Investigating the pairs we see that they are identified by the output value of the first command ($y$),
and that they determine whether the command on the parent or the child must precede the other.

In \cref{simple_reorder_equiv} we offer a formal proof that $\orderset{U}$ meets our expectations.
It is
an important property of simple sequences: namely, that
all of their valid reorderings are equivalent,
and all equivalent sequences are mere reorderings.
To prove this, first we prove the following:

%% TODO Define set operations on sequences?
% \noindent As a specific case, we can regard a simple sequence $S$
% as a set, and use the above algorithm to generate all possible orderings
% of the commands it contains.

@@

\begin{mylem}\label{simple_reorder_equiv}
For a simple sequence $S\nequiv\cbrk$,
\[ \orderset{S} = [S]_\equiv \cap \setssmp, \]
where $[S]_\equiv \cap \setssmp$ is the set of simple sequences equivalent to $S$.
\end{mylem}
\begin{proof}
First, as $S$ only has at most one command on each node,
any sequence in $\orderset{S}$ either breaks a filesystem,
or it leaves it in the same state as $S$.

TODO:
Second, based on the algorithm we also know that $\orderset{S}$ contains
all possible orderings of $S$ as the only ordering between commands
it prescribes are when the commands in the reverse order would break every filesystem.

### 2016-10-16 p2/works-inverse.tex
New ordering method; hyperlinks

In a sense, $A\worksmeqsign B$ is a weaker counterpart of $A\eqext B$, as the former
only requires that $B$ works where $A$ works, where the latter also requires
that their effect is the same.
It is easy to see that the following corollaries are true:

@@

\begin{mycor}\label{workschained}
% An inference rule in the algebra
$\worksmeqsign$ can be chained:
if $\sqs{B'}\subseteq\sqs{B}$, then
$\worksm{\sqs{A}}{\sqs{B}} \wedge \worksm{\sqs{B'}}{\sqs{C}} \Rightarrow \worksm{\sqs{A}}{\sqs{C}}$.
\end{mycor}

@@

\begin{myax}\label{combine_independent_commands}
The combination of independent commands works wherever the original commands work:
\[ \alpha\indep \beta \Rightarrow \worksm{\{\alpha, \beta\}}{\{\alpha\cc \beta\}}. \]
\end{myax}
\begin{proof}
We name this proposition a \namecref{combine_independent_commands} 
because to prove it, we must reach back to our filesystem model.
Let $\alpha=\cxynv$ and $\beta=\czwmv$.
We prove the proposition by contradiction and
assume that there is a filesystem $\FS$ for which
$\cxynv\aFS\neq\fsbroken$ and $\czwmv\aFS\neq\fsbroken$, but
$\cxynv\cc \czwmv\aFS=\fsbroken$.
We know $\cxynv\aFS\neq\fsbroken$ so it must be applying 
$\czwmv$ that breaks it.
Applying a command can only result in a broken filesystem in two cases.
First, if the input type does not match the filesystem,
but we know $[\FS(m)]=[w]$ and so
$[(\cxynv\aFS)(m)]=[w]$ as based on \cref{incomparable_is_independent}, $n\neq m$.
Second, if the new filesystem violates the tree property.
This again cannot be the case because we also know that $n\unrel m$
and the tree property only depends on the types of values at the parent and children of $m$,
which therefore cannot be changed by $\cxynv$.
\end{proof}

@@

\begin{mylem}\label{combine_independent_sequences}
The combination of independent sequences works wherever the original sequences work:
\[ S\indep T \Rightarrow \worksm{\{S,T\}}{\{S\cc T\}}. \]
\end{mylem}
\begin{proof}
Assume that there is a filesystem $\FS$ so that
$S\aFS\neq\fsbroken$ and $T\aFS\neq\fsbroken$, but
$(S\cc T)\FS=\fsbroken$.

@@

\begin{mylem}\label{indep_prefix_combine}
The combination of sequences with a common head and independent tails 
continues to work under the same conditions:
\begin{gather*}
 \worksmxb{\sqs{S}}{A\cc B} \,\wedge\, \worksmxb{\sqs{S}}{A\cc C} \,\wedge\, B\indep C \\
 \Longrightarrow \worksmxb{\sqs{S}}{A\cc B\cc C}
\end{gather*}
\end{mylem}
\begin{proof}
Based on \cref{repr_works_is_same} we have
$\worksmxb{\sqs{S}\T}{A\T\cc B\T}$ and $\worksmxb{\sqs{S}\T}{A\T\cc C\T}$,
and from \cref{r_invmove} we know
$\worksmxb{{A\T}^{-1}\cc \sqs{S\T}}{B\T}$ 
and $\worksmxb{{A\T}^{-1}\cc \sqs{S\T}}{C\T}$.
Using \cref{combine_independent_sequences} we get
$\worksmbb{B\T,C\T}{B\T\cc C\T}$,
and chaining these using \cref{workschained}
we get
$\worksmxb{{A\T}^{-1}\cc \sqs{S}\T}{B\T\cc C\T}$.

>>>>

In a sense, $A\worksmeqsign B$ is a weaker counterpart of $A\eqext B$, as the former
only requires that $B$ works where $A$ works, where the latter also requires
that where they work, their effect is the same.
It is easy to see that the following corollary is true:

@@

@@

\begin{myax}\label{combine_independent_commands}
The combination of independent commands works wherever both of the commands work:
\[ \alpha\indep \beta \Rightarrow \worksm{\{\alpha, \beta\}}{\{\alpha\cc \beta\}}. \]
\end{myax}
\begin{proof}
We name this proposition a \namecref{combine_independent_commands} 
because to prove it, we must reach back to our filesystem model.
Let $\alpha=\cxynv$ and $\beta=\czwmv$.
We prove the proposition by contradiction and
assume that there is a filesystem $\FS$ for which
$\cxynv\aFS\neq\fsbroken$ and $\czwmv\aFS\neq\fsbroken$, but
$\cxynv\cc \czwmv\aFS=\fsbroken$.
We know $\cxynv\aFS\neq\fsbroken$ so it must be applying 
$\czwmv$ that breaks it.
Applying a command can only result in a broken filesystem in two cases.
First, if the input type does not match the filesystem,
but we know $[\FS(m)]=[w]$ and so
$[(\cxynv\aFS)(m)]=[w]$ as based on \cref{incomparable_is_independent}, $n\neq m$.
Second, if the new filesystem violates the tree property.
This again cannot be the case because we also know that $n\unrel m$
and the tree property only depends on the types of values at the parent and children of $m$,
which therefore cannot be changed by $\cxynv$.
\end{proof}

@@

\begin{mylem}\label{combine_independent_sequences}
The combination of independent sequences works wherever both of the sequences work:
\[ S\indep T \Rightarrow \worksm{\{S,T\}}{\{S\cc T\}}. \]
\end{mylem}
\begin{proof}
Assume that there is a filesystem $\FS$ so that
$S\aFS\neq\fsbroken$ and $T\aFS\neq\fsbroken$, but
$(S\cc T)\FS=\fsbroken$.

@@

\begin{mylem}\label{indep_prefix_combine}
The combination of sequences with a common head and independent tails 
continues to work under the same conditions:
\begin{gather*}
 \worksmxb{\sqs{S}}{A\cc B} \,\wedge\, \worksmxb{\sqs{S}}{A\cc C} \,\wedge\, B\indep C \\
 \Longrightarrow \worksmxb{\sqs{S}}{A\cc B\cc C}
\end{gather*}
\end{mylem}
\begin{proof}
Based on \cref{repr_works_is_same} we have
$\worksmxb{\sqs{S}\T}{A\T\cc B\T}$ and $\worksmxb{\sqs{S}\T}{A\T\cc C\T}$,
and from \cref{r_invmove} we know
$\worksmxb{{A\T}^{-1}\cc \sqs{S\T}}{B\T}$ 
and $\worksmxb{{A\T}^{-1}\cc \sqs{S\T}}{C\T}$.
Using \cref{combine_independent_sequences} we get
$\worksmbb{B\T,C\T}{B\T\cc C\T}$,
and chaining these (as $\worksmeqsign$ is transitive)
we get
$\worksmxb{{A\T}^{-1}\cc \sqs{S}\T}{B\T\cc C\T}$.
